---
title: "Run CoPro on brain MERFISH data"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Run CoPro on brain MERFISH data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Content

This Vignette contains an example to run CoPro on a brain MERFISH data. 
The data is from Zhang et al. *Nature* 2023 (see reference 1). 
Here, we focus on detecting the co-progression between D1 and D2 neurons, as reported in the original publication.
 Here, we aim to show the unsupervised detection of continuous and coordinated cell progressions and how that is related to the tissue structure.

# load packages

```{r message=FALSE, warning=FALSE, results="hide"}
library("CoPro")
library_paths <- c(
  "fields", "ggplot2", "tictoc", "irlba",
  "Matrix", "ggsci", "ape"
)
# Load required libraries
lapply(library_paths, library, character.only = TRUE)
```

# set some parameters

We can change the number of PCs. For this dataset, this parameter does not affect the results much

```{r }
n_pca <- 40 ## number of top PCs to retain
cell_types <- c("061 STR D1 Gaba", "062 STR D2 Gaba")
n_cell_types <- length(cell_types)

horizontal_dist_scale <- 1
vertical_dist_scale <- 1
```

# Load data

Here we load the normalized gene matrix as well as the meta data. The cell type label as well as cell location information are in the meta data.

```{r, eval=FALSE}
ra <- readRDS("/Users/zhenmiao/Dropbox/Zhuang-ABCA-1.054_1/Zhuang_ABCA_1.054_subset_data.rds")
meta <- readRDS("/Users/zhenmiao/Dropbox/Zhuang-ABCA-1.054_1/Zhuang_ABCA_1.054_subset_metadata.rds")
```

## subset data

```{r}
subset_cells <- meta$subclass %in% cell_types

ra <- ra[subset_cells, ]
meta <- meta[subset_cells, ]
rownames(meta) <- meta$cell_label
## make sure coordinates are numeric
meta$x <- as.numeric(meta$x)
meta$y <- as.numeric(meta$y)
```

## show the in situ plot

```{r}
ggplot(data = meta) +
  geom_point(aes(x = x, y = y, color = subclass)) +
  theme_minimal() +
  coord_fixed(ratio = 1)
```

# load data into object

```{r create_object}
location_data <- meta[, c("x", "y")]
cell_type_data <- meta[, "subclass"]
br <- newCoProSingle(
  normalizedData = ra,
  locationData = location_data,
  metaData = meta,
  cellTypes = cell_type_data
)
br <- subsetData(br, cellTypesOfInterest = cell_types)
```

# compute distances

```{r distCompute}
br <- computePCA(br, nPCA = n_pca, center = TRUE, scale. = TRUE)
# sigma_square_choice <- c(0.001, 0.002, 0.005, 0.01, 0.02, 0.05, 0.1, 0.2)
sigma_value_choice <- c(0.1, 0.14, 0.2, 0.5)
## By default, we normalize the distance
br <- computeDistance(br, distType = "Euclidean2D", normalizeDistance = FALSE)
br <- computeKernelMatrix(br, sigmaValues = sigma_value_choice)
```

# run skrCCA and obtain correlations

```{r skrCCA}
br <- runSkrCCA(br, scalePCs = TRUE, maxIter = 500)
br <- computeNormalizedCorrelation(br)
```

# compute gene and cell scores

```{r geneCellScores}
br <- computeGeneAndCellScores(br)
```

# Look across different sigma values

Here, we hope to pick the sigma value with the highest normalized correlation value. We visualize the trend using ggplot2

```{r sigma_choice}
ncorr <- getNormCorr(br)

ggplot(data = ncorr, aes(
  x = sigmaValues,
  y = normalizedCorrelation, group = 1
)) +
  geom_point() +
  geom_line() +
  facet_wrap(vars(ct12,CC_index)) +
  xlab("Sigma squared") +
  ylab("Norm. Corr.") +
  ggtitle(label = "Norm. Corr. across sigma squared values") +
  theme_minimal()
```

From the figure above, the highest correlation is when sigma^2^ = 0.02, so we use this sigma value for the downstream analyses

# Cell scores correlation

We can visualize the cell scores with respect to one another by plotting X~A~K~AB~ against X~B~, where X~A~ represents the cell scores of each cell in cell type A, and X~B~ represents the cell scores of each cell in cell type B, and K~AB~ represents the Kernel matrix between cell type A and B.

```{r }
cell_A <- "061 STR D1 Gaba"
cell_B <- "062 STR D2 Gaba"

df_crtt <- getCorrTwoTypes(br,
  sigmaValueChoice = 0.14,
  cellTypeA = cell_A,
  cellTypeB = cell_B
)

ggplot(df_crtt) +
  geom_point(aes(x = AK, y = B)) +
  ggtitle(paste("Correlation plot between", cell_A, "and", cell_B)) +
  xlab(paste(cell_A, "%*% Kernal_AB")) +
  ylab(cell_B) +
  theme_minimal()

```

# Cell scores in situ

We can retrive the `data.frame` that contains the cell scores using the following function:

```{r}
br_cs <- getCellScoresInSitu(br,
  sigmaValueChoice = 0.14
)
```

We can visualize the cell scores in situ using ggplot.

```{r}
ggplot(data = br_cs) +
  geom_point(aes(
    x = x,
    y = y,
    color = cellScores_b
  ), size = 0.8) +
  scale_color_discrete(
    type =
      c("darkgreen", "#d4f8d4", "darkred", "#ffd8d8")
  ) +
  coord_fixed() +
  theme_minimal()
```

We can also visualize the continuous scores

```{r}
ggplot(data = br_cs) +
  geom_point(aes(
    x = x,
    y = y,
    color = cellScores
  ), size = 0.8) +
  coord_fixed() +
  theme_minimal()
```

# Permutation to establish null
```{r}
br <- runSkrCCAPermu(br, nPermu = 5L, permu_method = "bin",
                     num_bins_x = 10, num_bins_y = 10)

br <- computeNormalizedCorrelationPermu(br, tol = 1e-3)

nc_permu <- br@normalizedCorrelationPermu
# names(nc_permu) <- NULL
nc_permu <- do.call(rbind, nc_permu)
print(nc_permu)
```


# References:

The MERFISH data were downloaded from the following manuscript:

Zhang, M., Pan, X., Jung, W. *et al.* Molecularly defined and spatially resolved cell atlas of the whole mouse brain. *Nature* 624, 343–354 (2023). <https://doi.org/10.1038/s41586-023-06808-9>
