---
title: "Cross-cell-type co-progression (Brain MERFISH)"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Cross-cell-type co-progression (Brain MERFISH)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,
                      fig.width = 6, fig.height = 5)
```

## Overview

This vignette demonstrates **cross-cell-type co-progression** using brain
MERFISH data from Zhang et al. *Nature* 2023. We analyze D1 and D2
GABAergic neurons in the striatum, detecting coordinated spatial gene
expression patterns between these two cell populations.

## Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(CoPro)
library(ggplot2)
```

## Download and load data

```{r download-data}
data_path <- copro_download_data("brain_merfish")
dat <- readRDS(data_path)
```

## Visualize spatial layout

```{r plot-layout}
ggplot(dat$metaData) +
  geom_point(aes(x = x, y = y, color = subclass), size = 0.5) +
  coord_fixed() +
  theme_minimal() +
  ggtitle("D1 and D2 neurons in the striatum")
```

## Create CoPro object and run pipeline

```{r run-pipeline}
cell_types <- c("061 STR D1 Gaba", "062 STR D2 Gaba")

obj <- newCoProSingle(
  normalizedData = dat$normalizedData,
  locationData = dat$locationData,
  metaData = dat$metaData,
  cellTypes = dat$cellTypes
)
obj <- subsetData(obj, cellTypesOfInterest = cell_types)

# Core pipeline
obj <- computePCA(obj, nPCA = 40, center = TRUE, scale. = TRUE)
obj <- computeDistance(obj, distType = "Euclidean2D",
                       normalizeDistance = FALSE)
obj <- computeKernelMatrix(obj, sigmaValues = c(0.1, 0.14, 0.2, 0.5))
obj <- runSkrCCA(obj, scalePCs = TRUE, maxIter = 500)
obj <- computeNormalizedCorrelation(obj)
obj <- computeGeneAndCellScores(obj)
```

## Select optimal sigma

```{r plot-ncorr}
ncorr <- getNormCorr(obj)

ggplot(ncorr, aes(x = sigmaValues, y = normalizedCorrelation)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ ct12 + CC_index) +
  xlab("Sigma") +
  ylab("Normalized Correlation") +
  ggtitle("Normalized correlation across sigma values") +
  theme_minimal()
```

## Cross-type correlation

Visualize how cell scores in D1 neurons (smoothed by the spatial kernel)
correlate with D2 neuron scores:

```{r plot-cross-corr}
df_corr <- getCorrTwoTypes(obj,
  sigmaValueChoice = 0.14,
  cellTypeA = "061 STR D1 Gaba",
  cellTypeB = "062 STR D2 Gaba"
)

ggplot(df_corr) +
  geom_point(aes(x = AK, y = B), size = 0.5, alpha = 0.5) +
  xlab("D1 score (spatially smoothed)") +
  ylab("D2 score") +
  ggtitle("Cross-type correlation: D1 vs D2") +
  theme_minimal()
```

## In situ visualization

```{r plot-insitu}
cs <- getCellScoresInSitu(obj, sigmaValueChoice = 0.14)

ggplot(cs) +
  geom_point(aes(x = x, y = y, color = cellScores_b), size = 0.8) +
  scale_color_manual(values = c("darkgreen", "#d4f8d4",
                                 "darkred", "#ffd8d8")) +
  coord_fixed() +
  ggtitle("Cell scores in situ (D1 and D2 neurons)") +
  theme_minimal()

# Continuous scores
ggplot(cs) +
  geom_point(aes(x = x, y = y, color = cellScores), size = 0.8) +
  scale_color_viridis_c() +
  coord_fixed() +
  ggtitle("Continuous cell scores in situ") +
  theme_minimal()
```

## Permutation test

Establish statistical significance with spatial permutations:

```{r permutation}
obj <- runSkrCCAPermu(obj, nPermu = 5L, permu_method = "bin",
                       num_bins_x = 10, num_bins_y = 10)
obj <- computeNormalizedCorrelationPermu(obj, tol = 1e-3)

nc_permu <- do.call(rbind, obj@normalizedCorrelationPermu)
print(nc_permu)
```

## References

Zhang, M., Pan, X., Jung, W. *et al.* Molecularly defined and spatially
resolved cell atlas of the whole mouse brain. *Nature* 624, 343--354
(2023). <https://doi.org/10.1038/s41586-023-06808-9>

## Session info

```{r session-info}
sessionInfo()
```
