---
title: "Cross-cell-type co-progression with orthogonal axes (Colon Day 3)"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Cross-cell-type co-progression with orthogonal axes (Colon Day 3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,
                      fig.width = 7, fig.height = 5)
```

## Overview

This vignette demonstrates how CoPro detects **cross-cell-type
co-progression** using colon organoid Day 3 data. CoPro identifies
multiple orthogonal canonical components (CCs), each capturing an
independent axis of coordinated spatial variation between cell types.

In this dataset, we analyze three cell types---Epithelial, Fibroblast,
and Immune---and show that CC1 and CC2 capture distinct, biologically
meaningful spatial programs. We also compare CoPro axes with
independently derived **mucosal neighborhood (MU) labels** to validate
that the unsupervised axes recover known tissue structure.

## Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(CoPro)
library(ggplot2)
```

## Download and load data

```{r download-data}
data_path <- copro_download_data("colon_d3")
dat <- readRDS(data_path)

str(dat[c("normalizedData", "locationData", "cellTypes")], max.level = 1)
```

## Visualize the tissue

```{r plot-layout, fig.width=7, fig.height=6}
plot_df <- data.frame(
  x = dat$locationData$x,
  y = dat$locationData$y,
  celltype = dat$cellTypes
)

ggplot(plot_df, aes(x = x, y = y, color = celltype)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = c("Epithelial" = "#E41A1C",
                                 "Fibroblast" = "#377EB8",
                                 "Immune" = "#4DAF4A")) +
  coord_fixed() +
  ggtitle("Colon Day 3 organoid") +
  theme_classic() +
  theme(legend.position = "bottom")
```

### Mucosal neighborhood (MU) labels

The tissue has been independently annotated with mucosal neighborhood
labels (MU1--MU4) via Leiden clustering on spatial neighborhoods. MU4
represents an inflammation-associated microenvironment distinct from
MU1--3 (crypt-associated neighborhoods):

```{r plot-mu, fig.width=7, fig.height=6}
mu_df <- data.frame(
  x = dat$locationData$x,
  y = dat$locationData$y,
  MU = dat$metaData$Leiden_neigh
)

# Group MU1-3 together as in the manuscript
mu_df$MU_grouped <- ifelse(mu_df$MU %in% c("MU1", "MU2", "MU3"),
                            "MU1-3", as.character(mu_df$MU))
mu_df$MU_grouped <- factor(mu_df$MU_grouped, levels = c("MU1-3", "MU4"))

mu_colors <- c("MU1-3" = "#5c9137", "MU4" = "#614b83")

ggplot(mu_df, aes(x = x, y = y, color = MU_grouped)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = mu_colors) +
  coord_fixed() +
  ggtitle("Mucosal neighborhood (MU) labels") +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.title = element_blank())
```

## Create CoPro object

```{r create-object}
obj <- newCoProSingle(
  normalizedData = dat$normalizedData,
  locationData = dat$locationData,
  metaData = dat$metaData,
  cellTypes = dat$cellTypes
)

cell_types <- c("Epithelial", "Fibroblast", "Immune")
obj <- subsetData(obj, cellTypesOfInterest = cell_types)
```

## Run the CoPro pipeline

```{r run-pipeline}
# PCA
obj <- computePCA(obj, nPCA = 40, center = TRUE, scale. = TRUE)

# Distance and kernel
obj <- computeDistance(obj, distType = "Euclidean2D")

sigma_choice <- c(0.005, 0.01, 0.02, 0.05, 0.1)
obj <- computeKernelMatrix(obj, sigmaValues = sigma_choice)

# Sparse kernel CCA -- request 4 CCs to capture multiple axes
obj <- runSkrCCA(obj, scalePCs = TRUE, maxIter = 500, nCC = 4)

# Normalized correlation and scores
obj <- computeNormalizedCorrelation(obj)
obj <- computeGeneAndCellScores(obj)
```

## Select optimal sigma

```{r plot-ncorr, fig.width=8, fig.height=5}
ncorr <- getNormCorr(obj)

ggplot(ncorr, aes(x = sigmaValues, y = normalizedCorrelation)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ ct12 + CC_index, scales = "free_y") +
  xlab("Sigma") +
  ylab("Normalized Correlation") +
  ggtitle("Normalized correlation across sigma values") +
  theme_minimal()
```

## Orthogonal axes: CC1 vs CC2

A key feature of CoPro is that it extracts **multiple orthogonal axes**.
Each CC captures a distinct pattern of spatial co-variation:

### CC1 cell scores in situ

```{r insitu-cc1, fig.width=7, fig.height=6}
sigma_opt <- 0.01  # adjust based on ncorr plot

cs_cc1 <- getCellScoresInSitu(obj, sigmaValueChoice = sigma_opt,
                               ccIndex = 1)

ggplot(cs_cc1) +
  geom_point(aes(x = x, y = y, color = cellScores), size = 0.8) +
  scale_color_viridis_c() +
  facet_wrap(~ cellTypesSub) +
  coord_fixed() +
  ggtitle("CC1 cell scores in situ (by cell type)") +
  theme_minimal()
```

### CC2 cell scores in situ

```{r insitu-cc2, fig.width=7, fig.height=6}
cs_cc2 <- getCellScoresInSitu(obj, sigmaValueChoice = sigma_opt,
                               ccIndex = 2)

ggplot(cs_cc2) +
  geom_point(aes(x = x, y = y, color = cellScores), size = 0.8) +
  scale_color_viridis_c(option = "C") +
  facet_wrap(~ cellTypesSub) +
  coord_fixed() +
  ggtitle("CC2 cell scores in situ (by cell type)") +
  theme_minimal()
```

CC1 and CC2 capture distinct spatial gradients. Because CCA axes are
orthogonal, these represent independent programs of coordinated gene
expression across cell types.

## CoPro axes recover mucosal neighborhood structure

A key validation is that CoPro's unsupervised axes align with the
independently derived MU labels. We plot CC1 vs CC2 cell scores colored
by MU label, showing that MU4 (inflammation) occupies a distinct region
of the CoPro score space:

```{r cc1-vs-cc2-mu, fig.width=6, fig.height=4}
lmeta <- obj@metaDataSub
lmeta$cc1 <- lmeta[, paste0("cellScore_sigma_", sigma_opt, "_cc_index_1")]
lmeta$cc2 <- lmeta[, paste0("cellScore_sigma_", sigma_opt, "_cc_index_2")]
lmeta$MU_grouped <- ifelse(lmeta$Leiden_neigh == "MU4", "MU4", "MU1-3")

ggplot() +
  geom_point(data = lmeta, aes(x = cc1, y = cc2, color = MU_grouped),
             size = 0.1, alpha = 0.1) +
  geom_density_2d(data = lmeta[lmeta$MU_grouped == "MU1-3", ],
                  aes(x = cc1, y = cc2), color = "#5c9137",
                  linewidth = 0.5, alpha = 0.9) +
  geom_density_2d(data = lmeta[lmeta$MU_grouped == "MU4", ],
                  aes(x = cc1, y = cc2), color = "#614b83",
                  linewidth = 0.5, alpha = 0.9) +
  scale_color_manual(values = c("MU1-3" = "#5c9137", "MU4" = "#614b83")) +
  xlab("CoPro CC1") +
  ylab("CoPro CC2") +
  labs(color = "MU Label") +
  ggtitle("CoPro CC1 vs CC2 colored by MU label") +
  theme_classic()
```

## Example gene expression in situ

Top CoPro genes show spatially coherent expression patterns. Here we
visualize two marker genes: **Egr1** (immediate early gene, enriched in
the inflammation/CC2 axis) and **Mki67** (proliferation marker, enriched
in crypt base/CC1 axis):

```{r gene-insitu-egr1, fig.width=5, fig.height=4}
expr_df <- data.frame(
  x = dat$locationData$x,
  y = dat$locationData$y,
  Egr1 = dat$normalizedData[, "Egr1"],
  Mki67 = dat$normalizedData[, "Mki67"]
)

# Egr1 (inflammation marker)
expr_df <- expr_df[order(expr_df$Egr1), ]
ggplot(expr_df, aes(x = x, y = y, color = Egr1)) +
  geom_point(size = 0.2) +
  scale_color_gradientn(colors = c("gray90", "darkred")) +
  coord_fixed() +
  ggtitle("Egr1 (inflammation marker)") +
  theme_classic() +
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank())
```

```{r gene-insitu-mki67, fig.width=5, fig.height=4}
# Mki67 (proliferation marker)
expr_df <- expr_df[order(expr_df$Mki67), ]
ggplot(expr_df, aes(x = x, y = y, color = Mki67)) +
  geom_point(size = 0.2) +
  scale_color_gradientn(colors = c("gray90", "darkblue")) +
  coord_fixed() +
  ggtitle("Mki67 (proliferation marker)") +
  theme_classic() +
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank())
```

## Cross-type correlation plots

The correlation between cell types shows how strongly their spatial
programs are coupled:

```{r cross-corr, fig.width=6, fig.height=5}
# CC1: Epithelial vs Fibroblast
df_cc1 <- getCorrTwoTypes(obj,
  sigmaValueChoice = sigma_opt,
  cellTypeA = "Epithelial",
  cellTypeB = "Fibroblast",
  ccIndex = 1
)

ggplot(df_cc1) +
  geom_point(aes(x = AK, y = B), size = 0.5, alpha = 0.4) +
  xlab("Epithelial CC1 (spatially smoothed)") +
  ylab("Fibroblast CC1") +
  ggtitle("CC1: Epithelial-Fibroblast co-progression") +
  theme_minimal()

# CC2: Epithelial vs Fibroblast
df_cc2 <- getCorrTwoTypes(obj,
  sigmaValueChoice = sigma_opt,
  cellTypeA = "Epithelial",
  cellTypeB = "Fibroblast",
  ccIndex = 2
)

ggplot(df_cc2) +
  geom_point(aes(x = AK, y = B), size = 0.5, alpha = 0.4) +
  xlab("Epithelial CC2 (spatially smoothed)") +
  ylab("Fibroblast CC2") +
  ggtitle("CC2: Epithelial-Fibroblast co-progression") +
  theme_minimal()
```

## Gene scores

Identify which genes drive each co-progression axis:

```{r gene-scores}
# Compute regression-based gene scores (recommended)
obj <- computeRegressionGeneScores(obj, sigma = sigma_opt)
```

### Visualize top genes for CC1

```{r top-genes-cc1, fig.width=5, fig.height=5}
# Extract regression gene scores for Epithelial CC1
key <- paste0("geneScores|sigma", sigma_opt, "|Epithelial")
gs_epi <- obj@geneScoresRegression[[key]]

# Top 20 genes by absolute weight for CC1
gs_cc1 <- gs_epi[, 1]
top_genes <- head(sort(abs(gs_cc1), decreasing = TRUE), 20)
top_df <- data.frame(
  gene = factor(names(top_genes), levels = rev(names(top_genes))),
  weight = gs_cc1[names(top_genes)]
)
top_df$direction <- ifelse(top_df$weight > 0, "positive", "negative")

ggplot(top_df, aes(x = gene, y = weight, fill = direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("positive" = "darkred",
                                "negative" = "lightpink")) +
  ggtitle("Top 20 Epithelial genes (CC1, regression)") +
  theme_classic() +
  theme(legend.position = "none")
```

### Visualize top genes for CC2

```{r top-genes-cc2, fig.width=5, fig.height=5}
# Top 20 genes by absolute weight for CC2
gs_cc2 <- gs_epi[, 2]
top_genes_cc2 <- head(sort(abs(gs_cc2), decreasing = TRUE), 20)
top_df_cc2 <- data.frame(
  gene = factor(names(top_genes_cc2), levels = rev(names(top_genes_cc2))),
  weight = gs_cc2[names(top_genes_cc2)]
)
top_df_cc2$direction <- ifelse(top_df_cc2$weight > 0, "positive", "negative")

ggplot(top_df_cc2, aes(x = gene, y = weight, fill = direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("positive" = "darkred",
                                "negative" = "lightpink")) +
  ggtitle("Top 20 Epithelial genes (CC2, regression)") +
  theme_classic() +
  theme(legend.position = "none")
```

## Session info

```{r session-info}
sessionInfo()
```
