---
title: "Multi-slide analysis and score transfer (Colon Day 9)"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Multi-slide analysis and score transfer (Colon Day 9)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,
                      fig.width = 7, fig.height = 5)
```

## Overview

This vignette demonstrates CoPro's **multi-slide analysis** and **score
transfer** capabilities using colon organoid Day 9 data. At Day 9,
the tissue exhibits severe inflammation with heterogeneous disease
progression across different regions. CoPro detects a **disease
progression axis** that correlates with independently derived mucosal
neighborhood (MU) labels---in particular, **MU8** marks the most
severely inflamed niche.

When multiple tissue slides are available, CoPro can:

1. Learn co-progression patterns from a **reference slide**
2. **Transfer** the learned gene weights to new slides
3. Compare co-progression patterns across biological replicates

## Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(CoPro)
library(ggplot2)
```

## Download and load data

```{r download-data}
data_path <- copro_download_data("colon_d9")
dat <- readRDS(data_path)

# Three slides are included
cat("Slides:", paste(dat$selectedSlides, collapse = ", "), "\n")
cat("Total cells:", nrow(dat$normalizedData), "\n")
table(dat$slideID)
```

## Visualize the tissue

### Cell types

```{r plot-celltype, fig.width=7, fig.height=6}
plot_df <- data.frame(
  x = dat$locationData$x,
  y = dat$locationData$y,
  celltype = dat$cellTypes,
  slide = dat$slideID
)

ggplot(plot_df[plot_df$slide == dat$selectedSlides[1], ],
       aes(x = x, y = y, color = celltype)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = c("Epithelial" = "#E41A1C",
                                 "Fibroblast" = "#377EB8",
                                 "Immune" = "#4DAF4A")) +
  coord_fixed() +
  ggtitle(paste("Colon Day 9 --", dat$selectedSlides[1])) +
  theme_classic() +
  theme(legend.position = "bottom")
```

### Mucosal neighborhood (MU) labels

The tissue has been independently annotated with mucosal neighborhood
labels via Leiden clustering. **MU8** marks the most severely inflamed
niche, while MU1--3 are grouped as relatively normal crypt neighborhoods:

```{r plot-mu, fig.width=7, fig.height=6}
mu_df <- data.frame(
  x = dat$locationData$x,
  y = dat$locationData$y,
  MU = dat$metaData$Leiden_neigh,
  slide = dat$slideID
)

# Group MU labels as in the manuscript
mu_df$MU_grouped <- as.character(mu_df$MU)
mu_df$MU_grouped[mu_df$MU %in% c("MU1", "MU11")] <- "MU1-3"
mu_df <- mu_df[mu_df$MU_grouped %in% c("MU1-3", "MU5", "MU6", "MU7", "MU8"), ]
mu_df$MU_grouped <- factor(mu_df$MU_grouped,
                            levels = c("MU1-3", "MU5", "MU6", "MU7", "MU8"))

mu_colors <- c("MU1-3" = "#66C2A5", "MU5" = "#E5C494",
               "MU6" = "#8DA0CB", "MU7" = "#FC8D62", "MU8" = "#E78AC3")

ggplot(mu_df[mu_df$slide == dat$selectedSlides[1], ],
       aes(x = x, y = y, color = MU_grouped)) +
  geom_point(size = 0.5) +
  scale_color_manual(values = mu_colors) +
  coord_fixed() +
  ggtitle("Mucosal neighborhoods (MU labels)") +
  theme_classic() +
  theme(legend.title = element_blank(), legend.position = "right") +
  guides(color = guide_legend(override.aes = list(size = 3)))
```

## Strategy: Reference + target slides

We use the first slide as the **reference** to learn co-progression
patterns, then transfer scores to the other two slides.

```{r split-slides}
ref_slide <- dat$selectedSlides[1]
tar_slides <- dat$selectedSlides[2:3]

# Indices for reference and target
ref_idx <- dat$slideID == ref_slide
tar1_idx <- dat$slideID == tar_slides[1]
tar2_idx <- dat$slideID == tar_slides[2]
```

## Step 1: Run CoPro on the reference slide

```{r run-reference}
cell_types <- c("Epithelial", "Fibroblast", "Immune")

# Create reference CoPro object
ref_obj <- newCoProSingle(
  normalizedData = dat$normalizedData[ref_idx, ],
  locationData = dat$locationData[ref_idx, ],
  metaData = dat$metaData[ref_idx, ],
  cellTypes = dat$cellTypes[ref_idx]
)
ref_obj <- subsetData(ref_obj, cellTypesOfInterest = cell_types)

# Run pipeline
ref_obj <- computePCA(ref_obj, nPCA = 40, center = TRUE, scale. = TRUE)
ref_obj <- computeDistance(ref_obj, distType = "Euclidean2D")

sigma_choice <- c(0.005, 0.01, 0.02, 0.05, 0.1)
ref_obj <- computeKernelMatrix(ref_obj, sigmaValues = sigma_choice)
ref_obj <- runSkrCCA(ref_obj, scalePCs = TRUE, maxIter = 500, nCC = 2)
ref_obj <- computeNormalizedCorrelation(ref_obj)
ref_obj <- computeGeneAndCellScores(ref_obj)
ref_obj <- computeRegressionGeneScores(ref_obj)
```

### Reference slide results

```{r ref-results, fig.width=7, fig.height=6}
sigma_opt <- 0.01  # adjust based on normalized correlation

cs_ref <- getCellScoresInSitu(ref_obj, sigmaValueChoice = sigma_opt)

ggplot(cs_ref) +
  geom_point(aes(x = x, y = y, color = cellScores), size = 0.8) +
  scale_color_viridis_c() +
  facet_wrap(~ cellTypesSub) +
  coord_fixed() +
  ggtitle(paste("Reference slide:", ref_slide)) +
  theme_minimal()
```

### Disease axis vs MU labels

The CoPro disease axis should correlate with MU labels, with **MU8**
cells showing the highest scores:

```{r mu-boxplot, fig.width=6, fig.height=4}
ref_meta <- ref_obj@metaDataSub
ref_meta$cell_score <- ref_meta[, paste0("cellScore_sigma_", sigma_opt,
                                          "_cc_index_1")]

# Assign grouped MU labels
ref_meta$MU_grouped <- as.character(ref_meta$Leiden_neigh)
ref_meta$MU_grouped[ref_meta$MU_grouped %in% c("MU1", "MU11")] <- "MU1-3"
ref_meta <- ref_meta[ref_meta$MU_grouped %in%
                      c("MU1-3", "MU5", "MU6", "MU7", "MU8"), ]
ref_meta$MU_grouped <- factor(ref_meta$MU_grouped,
                               levels = c("MU1-3", "MU5", "MU6", "MU7", "MU8"))

ggplot(ref_meta, aes(x = MU_grouped, y = cell_score, fill = MU_grouped)) +
  geom_boxplot(outlier.size = 0.3) +
  scale_fill_manual(values = mu_colors, guide = "none") +
  labs(x = "Mucosal Neighborhood", y = "CoPro disease score (CC1)",
       title = "Disease scores by MU label",
       subtitle = "MU8 = most inflamed niche") +
  theme_classic()
```

### Cell type proportions along the disease axis

As disease severity increases, the cell type composition shifts---immune
cell proportion increases while epithelial proportion decreases:

```{r celltype-prop, fig.width=6, fig.height=4}
ref_meta_all <- ref_obj@metaDataSub
ref_meta_all$cell_score <- ref_meta_all[, paste0("cellScore_sigma_",
                                                   sigma_opt, "_cc_index_1")]

# Bin cells by disease score
ref_meta_all$score_bin <- cut(ref_meta_all$cell_score, breaks = 20)

# Calculate proportions
prop_df <- do.call(rbind, lapply(split(ref_meta_all, ref_meta_all$score_bin),
  function(x) {
    data.frame(
      score_mid = mean(x$cell_score, na.rm = TRUE),
      Epithelial = mean(x$Tier1 == "Epithelial"),
      Fibroblast = mean(x$Tier1 == "Fibroblast"),
      Immune = mean(x$Tier1 == "Immune")
    )
  }
))
prop_long <- reshape(prop_df, direction = "long",
                     varying = c("Epithelial", "Fibroblast", "Immune"),
                     v.names = "proportion", timevar = "celltype",
                     times = c("Epithelial", "Fibroblast", "Immune"))

ggplot(prop_long, aes(x = score_mid, y = proportion, color = celltype)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  scale_color_manual(values = c("Epithelial" = "#E41A1C",
                                 "Fibroblast" = "#377EB8",
                                 "Immune" = "#4DAF4A")) +
  labs(x = "CoPro disease score (CC1)", y = "Cell type proportion",
       title = "Cell type composition along the disease axis") +
  theme_classic() +
  theme(legend.position = "bottom")
```

### Top genes associated with the disease axis

```{r top-genes, fig.width=5, fig.height=5}
# Immune genes on the disease axis
key_imm <- paste0("geneScores|sigma", sigma_opt, "|Immune")
gs_imm <- ref_obj@geneScoresRegression[[key_imm]]
gs_imm_cc1 <- gs_imm[, 1]

top_imm <- head(sort(abs(gs_imm_cc1), decreasing = TRUE), 20)
top_imm_df <- data.frame(
  gene = factor(names(top_imm), levels = rev(names(top_imm))),
  weight = gs_imm_cc1[names(top_imm)]
)
top_imm_df$direction <- ifelse(top_imm_df$weight > 0, "inflamed", "normal")

ggplot(top_imm_df, aes(x = gene, y = weight, fill = direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("inflamed" = "darkgreen",
                                "normal" = "lightgreen")) +
  ggtitle("Top 20 Immune genes (CC1, regression)") +
  theme_classic() +
  theme(legend.position = "bottom")
```

## Step 2: Create target CoPro objects

```{r create-targets}
# Target slide 1
tar1_obj <- newCoProSingle(
  normalizedData = dat$normalizedData[tar1_idx, ],
  locationData = dat$locationData[tar1_idx, ],
  metaData = dat$metaData[tar1_idx, ],
  cellTypes = dat$cellTypes[tar1_idx]
)
tar1_obj <- subsetData(tar1_obj, cellTypesOfInterest = cell_types)
tar1_obj <- computePCA(tar1_obj, nPCA = 40, center = TRUE, scale. = TRUE)
tar1_obj <- computeDistance(tar1_obj, distType = "Euclidean2D")
tar1_obj <- computeKernelMatrix(tar1_obj, sigmaValues = sigma_choice)

# Target slide 2
tar2_obj <- newCoProSingle(
  normalizedData = dat$normalizedData[tar2_idx, ],
  locationData = dat$locationData[tar2_idx, ],
  metaData = dat$metaData[tar2_idx, ],
  cellTypes = dat$cellTypes[tar2_idx]
)
tar2_obj <- subsetData(tar2_obj, cellTypesOfInterest = cell_types)
tar2_obj <- computePCA(tar2_obj, nPCA = 40, center = TRUE, scale. = TRUE)
tar2_obj <- computeDistance(tar2_obj, distType = "Euclidean2D")
tar2_obj <- computeKernelMatrix(tar2_obj, sigmaValues = sigma_choice)
```

## Step 3: Transfer scores

Transfer the learned gene weights from the reference to each target slide:

```{r transfer-scores}
# Transfer to target 1
tar1_scores <- getTransferCellScores(
  ref_obj = ref_obj,
  tar_obj = tar1_obj,
  sigma_choice = sigma_opt,
  gene_score_type = "regression"
)

# Transfer to target 2
tar2_scores <- getTransferCellScores(
  ref_obj = ref_obj,
  tar_obj = tar2_obj,
  sigma_choice = sigma_opt,
  gene_score_type = "regression"
)
```

## Step 4: Visualize transferred scores

```{r plot-transfer, fig.width=7, fig.height=10}
# Build data frames for target slides from transferred scores
build_transfer_df <- function(tar_dat, tar_scores, slide_name, sigma) {
  rows <- list()
  for (ct in cell_types) {
    ct_key <- paste0("geneScores|sigma", sigma, "|", ct)
    if (ct_key %in% names(tar_scores)) {
      ct_idx <- tar_dat$cellTypes == ct
      rows[[ct]] <- data.frame(
        x = tar_dat$locationData$x[ct_idx],
        y = tar_dat$locationData$y[ct_idx],
        cellScores = tar_scores[[ct_key]][, 1],
        cellTypesSub = ct,
        slide = slide_name
      )
    }
  }
  do.call(rbind, rows)
}

# Build target data using the original dat split by slide indices
tar1_dat <- list(
  locationData = dat$locationData[tar1_idx, ],
  cellTypes = dat$cellTypes[tar1_idx]
)
tar2_dat <- list(
  locationData = dat$locationData[tar2_idx, ],
  cellTypes = dat$cellTypes[tar2_idx]
)

tar1_cs <- build_transfer_df(tar1_dat, tar1_scores, tar_slides[1], sigma_opt)
tar2_cs <- build_transfer_df(tar2_dat, tar2_scores, tar_slides[2], sigma_opt)

# Add slide label to reference
cs_ref$slide <- ref_slide

all_cs <- rbind(
  cs_ref[, c("x", "y", "cellScores", "cellTypesSub", "slide")],
  tar1_cs,
  tar2_cs
)

ggplot(all_cs) +
  geom_point(aes(x = x, y = y, color = cellScores), size = 0.5) +
  scale_color_viridis_c() +
  facet_grid(slide ~ cellTypesSub) +
  coord_fixed() +
  ggtitle("CC1 scores: reference (top) vs transferred (bottom)") +
  theme_minimal() +
  theme(strip.text = element_text(size = 8))
```

## Step 5: Transferred scores vs MU labels on target slides

Check whether the transferred disease scores also separate MU labels.
Here we show the reference slide MU boxplot (already computed above)
alongside the MU labels in spatial context for a target slide:

```{r transfer-mu-spatial, fig.width=7, fig.height=6}
# Show MU labels spatially on target slide 1
tar1_mu <- data.frame(
  x = dat$locationData$x[tar1_idx],
  y = dat$locationData$y[tar1_idx],
  MU = dat$metaData$Leiden_neigh[tar1_idx]
)
tar1_mu$MU_grouped <- as.character(tar1_mu$MU)
tar1_mu$MU_grouped[tar1_mu$MU_grouped %in% c("MU1", "MU11")] <- "MU1-3"
tar1_mu <- tar1_mu[tar1_mu$MU_grouped %in%
                    c("MU1-3", "MU5", "MU6", "MU7", "MU8"), ]
tar1_mu$MU_grouped <- factor(tar1_mu$MU_grouped,
                              levels = c("MU1-3", "MU5", "MU6", "MU7", "MU8"))

ggplot(tar1_mu, aes(x = x, y = y, color = MU_grouped)) +
  geom_point(size = 0.5) +
  scale_color_manual(values = mu_colors) +
  coord_fixed() +
  ggtitle(paste("MU labels --", tar_slides[1])) +
  theme_classic() +
  theme(legend.title = element_blank())
```

## Step 6: Assess transfer consistency

Compare the normalized correlation between the reference and transferred
slides to assess whether the co-progression pattern is consistent:

```{r transfer-ncorr}
tar1_ncorr <- getTransferNormCorr(
  tar_obj = tar1_obj,
  transfer_cell_scores = tar1_scores,
  sigma_choice = sigma_opt
)

tar2_ncorr <- getTransferNormCorr(
  tar_obj = tar2_obj,
  transfer_cell_scores = tar2_scores,
  sigma_choice = sigma_opt
)

cat("Reference norm. corr.:\n")
ref_ncorr <- getNormCorr(ref_obj)
print(ref_ncorr[ref_ncorr$sigmaValues == sigma_opt, ])

cat("\nTarget 1 transferred norm. corr.:\n")
print(tar1_ncorr)

cat("\nTarget 2 transferred norm. corr.:\n")
print(tar2_ncorr)
```

High transferred normalized correlations indicate that the co-progression
pattern learned from the reference generalizes well to independent slides.

## Alternative: Multi-slide CoPro

For joint analysis across all slides simultaneously, use `newCoProMulti`:

```{r multi-slide}
multi_obj <- newCoProMulti(
  normalizedData = dat$normalizedData,
  locationData = dat$locationData,
  metaData = dat$metaData,
  cellTypes = dat$cellTypes,
  slideID = dat$slideID
)
multi_obj <- subsetData(multi_obj, cellTypesOfInterest = cell_types)

# The rest of the pipeline is identical
multi_obj <- computePCA(multi_obj, nPCA = 40, center = TRUE, scale. = TRUE)
multi_obj <- computeDistance(multi_obj, distType = "Euclidean2D")
multi_obj <- computeKernelMatrix(multi_obj, sigmaValues = sigma_choice)
multi_obj <- runSkrCCA(multi_obj, scalePCs = TRUE, maxIter = 500, nCC = 2)
multi_obj <- computeNormalizedCorrelation(multi_obj)
multi_obj <- computeGeneAndCellScores(multi_obj)
```

The multi-slide approach optimizes CCA weights jointly across all slides,
while the transfer approach trains on one slide and evaluates generalization.
Both are useful depending on your analytical question.

## Session info

```{r session-info}
sessionInfo()
```
