---
title: "Multi-slide joint analysis (Colon Day 3)"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Multi-slide joint analysis (Colon Day 3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,
                      fig.width = 7, fig.height = 5)
```

## Overview

When multiple tissue slides are available from the same condition, CoPro
can analyze them **jointly** using `newCoProMulti`. Joint analysis pools
spatial relationships across slides, yielding more robust co-progression
axes than single-slide analysis. This is especially valuable when
individual slides may capture only part of the tissue heterogeneity.

This vignette walks through the full multi-slide workflow using three
colon organoid Day 3 slides:

1. Create a `CoProMulti` object with `newCoProMulti`
2. Run the standard CoPro pipeline (PCA, kernel, CCA, scores)
3. Visualize per-slide results
4. Assess cross-slide consistency via score transfer
5. Validate against independently derived mucosal neighborhood labels

For a complementary example showing the **reference-plus-transfer**
workflow (train on one slide, transfer to others), see the
`colon_d9_multi_slide` vignette.

## Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(CoPro)
library(ggplot2)
```

## Download and load data

```{r download-data}
data_path <- copro_download_data("colon_d3_multi")
dat <- readRDS(data_path)

cat("Slides:", paste(dat$selectedSlides, collapse = ", "), "\n")
cat("Total cells:", nrow(dat$normalizedData), "\n")
cat("Genes:", ncol(dat$normalizedData), "\n")
table(dat$slideID)
```

## Visualize the tissue

### Cell types across slides

```{r plot-celltype, fig.width=9, fig.height=4}
plot_df <- data.frame(
  x = dat$locationData$x,
  y = dat$locationData$y,
  celltype = dat$cellTypes,
  slide = dat$slideID
)

# Short slide labels for plotting
slide_labels <- setNames(
  paste0("Slide ", seq_along(dat$selectedSlides)),
  dat$selectedSlides
)
plot_df$slide_label <- slide_labels[plot_df$slide]

ggplot(plot_df, aes(x = x, y = y, color = celltype)) +
  geom_point(size = 0.3, alpha = 0.6) +
  scale_color_manual(values = c("Epithelial" = "#E41A1C",
                                 "Fibroblast" = "#377EB8",
                                 "Immune" = "#4DAF4A")) +
  facet_wrap(~ slide_label) +
  coord_fixed() +
  ggtitle("Colon Day 3 -- three slides") +
  theme_classic() +
  theme(legend.position = "bottom")
```

### Mucosal neighborhood (MU) labels

Each slide has been independently annotated with mucosal neighborhood
labels. MU4 marks an inflammation-associated microenvironment:

```{r plot-mu, fig.width=9, fig.height=4}
mu_df <- data.frame(
  x = dat$locationData$x,
  y = dat$locationData$y,
  MU = dat$metaData$Leiden_neigh,
  slide = dat$slideID
)
mu_df$slide_label <- slide_labels[mu_df$slide]

mu_df$MU_grouped <- ifelse(mu_df$MU %in% c("MU1", "MU2", "MU3"),
                            "MU1-3", as.character(mu_df$MU))
mu_df <- mu_df[mu_df$MU_grouped %in% c("MU1-3", "MU4"), ]
mu_df$MU_grouped <- factor(mu_df$MU_grouped, levels = c("MU1-3", "MU4"))

mu_colors <- c("MU1-3" = "#5c9137", "MU4" = "#614b83")

ggplot(mu_df, aes(x = x, y = y, color = MU_grouped)) +
  geom_point(size = 0.3, alpha = 0.6) +
  scale_color_manual(values = mu_colors) +
  facet_wrap(~ slide_label) +
  coord_fixed() +
  ggtitle("Mucosal neighborhood labels across slides") +
  theme_classic() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

## Create a multi-slide CoPro object

`newCoProMulti` takes a `slideID` vector that tells CoPro which cells
belong to which slide. The pipeline then computes distance and kernel
matrices **within** each slide (cells on different slides are never
treated as spatial neighbors), while CCA optimizes a shared set of
weights across all slides jointly.

```{r create-object}
cell_types <- c("Epithelial", "Fibroblast", "Immune")

multi_obj <- newCoProMulti(
  normalizedData = dat$normalizedData,
  locationData = dat$locationData,
  metaData = dat$metaData,
  cellTypes = dat$cellTypes,
  slideID = dat$slideID
)
multi_obj <- subsetData(multi_obj, cellTypesOfInterest = cell_types)
```

## Run the CoPro pipeline

The pipeline steps are identical to single-slide analysis. Internally,
CoPro handles the multi-slide structure automatically---kernel matrices
are block-diagonal (no cross-slide spatial interactions), and CCA finds
weights that maximize the pooled canonical correlation.

```{r run-pipeline}
multi_obj <- computePCA(multi_obj, nPCA = 40, center = TRUE, scale. = TRUE)
multi_obj <- computeDistance(multi_obj, distType = "Euclidean2D")

sigma_choice <- c(0.005, 0.01, 0.02, 0.05, 0.1)
multi_obj <- computeKernelMatrix(multi_obj, sigmaValues = sigma_choice)

multi_obj <- runSkrCCA(multi_obj, scalePCs = TRUE, maxIter = 500, nCC = 4)
multi_obj <- computeNormalizedCorrelation(multi_obj)
multi_obj <- computeGeneAndCellScores(multi_obj)
```

## Select optimal sigma

```{r plot-ncorr, fig.width=8, fig.height=5}
ncorr <- getNormCorr(multi_obj)

ggplot(ncorr, aes(x = sigmaValues, y = normalizedCorrelation)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ ct12 + CC_index, scales = "free_y") +
  xlab("Sigma") +
  ylab("Normalized Correlation") +
  ggtitle("Normalized correlation across sigma values") +
  theme_minimal()
```

```{r select-sigma}
sigma_opt <- 0.01  # adjust based on the ncorr plot above
```

## Visualize cell scores per slide

A key advantage of multi-slide analysis: the same CCA weights produce
cell scores on every slide. Consistent spatial patterns across slides
indicate a robust biological signal.

### CC1 scores across all slides

```{r insitu-cc1, fig.width=9, fig.height=4}
cs <- getCellScoresInSitu(multi_obj, sigmaValueChoice = sigma_opt,
                           ccIndex = 1)

# Add slide labels
cs$slide <- multi_obj@metaDataSub$Slice_ID[
  match(rownames(cs), rownames(multi_obj@metaDataSub))
]
if (is.null(cs$slide)) {
  cs$slide <- multi_obj@metaDataSub[
    match(paste(cs$x, cs$y), paste(multi_obj@locationDataSub$x,
                                    multi_obj@locationDataSub$y)),
    "Slice_ID"
  ]
}
cs$slide_label <- slide_labels[cs$slide]

ggplot(cs, aes(x = x, y = y, color = cellScores)) +
  geom_point(size = 0.5) +
  scale_color_viridis_c() +
  facet_grid(slide_label ~ cellTypesSub) +
  coord_fixed() +
  ggtitle("CC1 cell scores across slides") +
  theme_minimal() +
  theme(strip.text = element_text(size = 8))
```

### CC2 scores across all slides

```{r insitu-cc2, fig.width=9, fig.height=4}
cs2 <- getCellScoresInSitu(multi_obj, sigmaValueChoice = sigma_opt,
                            ccIndex = 2)
cs2$slide <- cs$slide
cs2$slide_label <- cs$slide_label

ggplot(cs2, aes(x = x, y = y, color = cellScores)) +
  geom_point(size = 0.5) +
  scale_color_viridis_c(option = "C") +
  facet_grid(slide_label ~ cellTypesSub) +
  coord_fixed() +
  ggtitle("CC2 cell scores across slides") +
  theme_minimal() +
  theme(strip.text = element_text(size = 8))
```

## CoPro axes vs MU labels

The CoPro axes are learned in a fully unsupervised manner. As a
validation, we check whether the unsupervised CC1/CC2 axes separate the
independently derived MU labels. We plot CC1 vs CC2 for each slide,
colored by MU label:

```{r cc1-vs-cc2-mu, fig.width=9, fig.height=4}
lmeta <- multi_obj@metaDataSub
lmeta$cc1 <- lmeta[, paste0("cellScore_sigma_", sigma_opt, "_cc_index_1")]
lmeta$cc2 <- lmeta[, paste0("cellScore_sigma_", sigma_opt, "_cc_index_2")]
lmeta$MU_grouped <- ifelse(lmeta$Leiden_neigh %in% c("MU1", "MU2", "MU3"),
                            "MU1-3",
                            ifelse(lmeta$Leiden_neigh == "MU4", "MU4", NA))
lmeta <- lmeta[!is.na(lmeta$MU_grouped), ]
lmeta$slide_label <- slide_labels[lmeta$Slice_ID]

ggplot(lmeta, aes(x = cc1, y = cc2, color = MU_grouped)) +
  geom_point(size = 0.3, alpha = 0.3) +
  scale_color_manual(values = mu_colors) +
  facet_wrap(~ slide_label) +
  labs(x = "CoPro CC1", y = "CoPro CC2", color = "MU Label") +
  ggtitle("CoPro CC1 vs CC2 colored by MU label") +
  theme_classic() +
  theme(legend.position = "bottom")
```

MU4 (inflammation) occupies a distinct region of the CoPro score space
on each slide, confirming that the jointly learned axes capture
biologically meaningful structure that generalizes across slides.

## Cross-type correlation

```{r cross-corr, fig.width=6, fig.height=5}
df_cc1 <- getCorrTwoTypes(multi_obj,
  sigmaValueChoice = sigma_opt,
  cellTypeA = "Epithelial",
  cellTypeB = "Fibroblast",
  ccIndex = 1
)

ggplot(df_cc1) +
  geom_point(aes(x = AK, y = B), size = 0.3, alpha = 0.3) +
  xlab("Epithelial CC1 (spatially smoothed)") +
  ylab("Fibroblast CC1") +
  ggtitle("Epithelial-Fibroblast co-progression (CC1, all slides)") +
  theme_minimal()
```

## Gene scores

```{r gene-scores}
multi_obj <- computeRegressionGeneScores(multi_obj, sigma = sigma_opt)
```

### Top genes for CC1

```{r top-genes-cc1, fig.width=5, fig.height=5}
key <- paste0("geneScores|sigma", sigma_opt, "|Epithelial")
gs_epi <- multi_obj@geneScoresRegression[[key]]

gs_cc1 <- gs_epi[, 1]
top_genes <- head(sort(abs(gs_cc1), decreasing = TRUE), 20)
top_df <- data.frame(
  gene = factor(names(top_genes), levels = rev(names(top_genes))),
  weight = gs_cc1[names(top_genes)]
)
top_df$direction <- ifelse(top_df$weight > 0, "positive", "negative")

ggplot(top_df, aes(x = gene, y = weight, fill = direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("positive" = "darkred",
                                "negative" = "lightpink")) +
  ggtitle("Top 20 Epithelial genes (CC1, regression)") +
  theme_classic() +
  theme(legend.position = "none")
```

## Assessing cross-slide consistency via transfer

Even when using joint analysis, it is useful to verify that the learned
gene program transfers well to individual slides. Here we run CoPro on
one slide as the reference and transfer its gene weights to each of the
other slides, comparing the transferred scores with the jointly learned
scores.

```{r transfer-setup}
ref_slide <- dat$selectedSlides[1]
tar_slide <- dat$selectedSlides[2]

ref_idx <- dat$slideID == ref_slide
tar_idx <- dat$slideID == tar_slide
```

```{r transfer-ref}
ref_obj <- newCoProSingle(
  normalizedData = dat$normalizedData[ref_idx, ],
  locationData = dat$locationData[ref_idx, ],
  metaData = dat$metaData[ref_idx, ],
  cellTypes = dat$cellTypes[ref_idx]
)
ref_obj <- subsetData(ref_obj, cellTypesOfInterest = cell_types)
ref_obj <- computePCA(ref_obj, nPCA = 40, center = TRUE, scale. = TRUE)
ref_obj <- computeDistance(ref_obj, distType = "Euclidean2D")
ref_obj <- computeKernelMatrix(ref_obj, sigmaValues = sigma_choice)
ref_obj <- runSkrCCA(ref_obj, scalePCs = TRUE, maxIter = 500, nCC = 4)
ref_obj <- computeNormalizedCorrelation(ref_obj)
ref_obj <- computeGeneAndCellScores(ref_obj)
ref_obj <- computeRegressionGeneScores(ref_obj)
```

```{r transfer-tar}
tar_obj <- newCoProSingle(
  normalizedData = dat$normalizedData[tar_idx, ],
  locationData = dat$locationData[tar_idx, ],
  metaData = dat$metaData[tar_idx, ],
  cellTypes = dat$cellTypes[tar_idx]
)
tar_obj <- subsetData(tar_obj, cellTypesOfInterest = cell_types)
tar_obj <- computePCA(tar_obj, nPCA = 40, center = TRUE, scale. = TRUE)
tar_obj <- computeDistance(tar_obj, distType = "Euclidean2D")
tar_obj <- computeKernelMatrix(tar_obj, sigmaValues = sigma_choice)
```

```{r transfer-scores}
tar_scores <- getTransferCellScores(
  ref_obj = ref_obj,
  tar_obj = tar_obj,
  sigma_choice = sigma_opt,
  gene_score_type = "regression"
)
```

### Transferred normalized correlation

High transferred normalized correlation indicates that the co-progression
pattern learned from one slide generalizes to the other:

```{r transfer-ncorr}
tar_ncorr <- getTransferNormCorr(
  tar_obj = tar_obj,
  transfer_cell_scores = tar_scores,
  sigma_choice = sigma_opt
)

cat("Reference slide normalized correlation:\n")
ref_ncorr <- getNormCorr(ref_obj)
print(ref_ncorr[ref_ncorr$sigmaValues == sigma_opt, ])

cat("\nTransferred normalized correlation:\n")
print(tar_ncorr)
```

### Compare transferred vs jointly learned scores

```{r compare-scores, fig.width=6, fig.height=4}
# Jointly learned scores for the target slide (from multi_obj)
joint_meta <- multi_obj@metaDataSub
joint_meta_tar <- joint_meta[joint_meta$Slice_ID == tar_slide, ]
joint_scores <- joint_meta_tar[, paste0("cellScore_sigma_", sigma_opt,
                                         "_cc_index_1")]

# Transferred scores for the target slide (from ref -> tar transfer)
# Match by cell type to compare
for (ct in cell_types) {
  ct_key <- paste0("geneScores|sigma", sigma_opt, "|", ct)
  if (ct_key %in% names(tar_scores)) {
    ct_idx_tar <- tar_obj@cellTypesSub == ct
    ct_idx_joint <- joint_meta_tar$Tier1 == ct
    if (sum(ct_idx_tar) > 0 && sum(ct_idx_joint) > 0) {
      n_match <- min(sum(ct_idx_tar), sum(ct_idx_joint))
      r_val <- cor(
        tar_scores[[ct_key]][seq_len(n_match), 1],
        joint_scores[ct_idx_joint][seq_len(n_match)]
      )
      cat(sprintf("  %s: r = %.3f (n = %d cells)\n", ct, r_val, n_match))
    }
  }
}
```

High correlation between transferred and jointly learned scores
confirms that both approaches recover the same biological signal.

## Key points

1. **`newCoProMulti`** pools spatial information across slides, producing
   a single set of CCA weights. This is the recommended approach when
   all slides come from the same biological condition.
2. **Per-slide visualization** of jointly learned scores reveals whether
   the co-progression pattern is consistent across tissue replicates.
3. **Score transfer** provides an independent consistency check: weights
   learned on one slide should produce similar cell scores on another.
4. The pipeline steps (`computePCA`, `computeDistance`, etc.) are the
   same for single-slide and multi-slide objects---only the constructor
   (`newCoProMulti` vs `newCoProSingle`) differs.
5. For a **transfer-focused** workflow (train on a reference, predict on
   targets), see the `colon_d9_multi_slide` vignette.

## Session info

```{r session-info}
sessionInfo()
```
