---
title: "Handling very large datasets (Xenium, large MERFISH)"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Handling very large datasets (Xenium, large MERFISH)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# This is a how-to guide: the code shows the recommended calls for
# datasets that are too large to ship in a vignette, so chunks are not
# executed. Every call below is a valid, current CoPro API call.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

Imaging-based platforms such as **10x Xenium** and large **MERFISH** panels
routinely produce hundreds of thousands to millions of cells per slide. Two
things blow up with the number of cells `n`:

1. **The expression matrix** (`n` cells x genes).
2. **The spatial distance / kernel matrix**, which is `n x n` per
   cell-type block if built densely — the dominant cost at scale.

CoPro handles each with a memory-efficient path. This guide collects the
settings that keep a large run within memory while producing results that are
numerically identical to the classic dense pipeline. The scientific
interpretation of the output (cell scores, gene scores, normalized
correlation) is unchanged; only how they are computed differs.

The single most important change for large data is: **do not call
`computeDistance()` — use the sparse kernel path instead.**

## 1. Store expression sparsely or on disk

`newCoProSingle()` and `newCoProMulti()` accept three matrix types for
`normalizedData`:

- a base `matrix` (dense),
- a **sparse** `Matrix::dgCMatrix` (recommended for imaging panels, which are
  mostly zeros), or
- a **BPCells** `IterableMatrix`, which keeps the counts on disk and streams
  them (recommended when the matrix does not fit in RAM).

```{r input-types}
library(CoPro)

# Sparse in-memory matrix (genes are columns, cells are rows):
obj <- newCoProSingle(
  normalizedData = sparse_expr,     # a cells x genes dgCMatrix
  locationData   = loc_df,          # data.frame with x, y (and optional z)
  metaData       = meta_df,
  cellTypes      = cell_type_labels
)

# On-disk with BPCells (install with remotes::install_github("bnprks/BPCells/r")):
# `bp_mat` is a BPCells IterableMatrix; computePCA() below will call
# BPCells::svds() so the PCA never materializes a dense matrix.
obj <- newCoProSingle(
  normalizedData = bp_mat,
  locationData   = loc_df,
  metaData       = meta_df,
  cellTypes      = cell_type_labels
)
```

## 2. Subset to the cell types of interest early

A Xenium slide may contain dozens of cell types, but a co-progression analysis
usually targets two (cross-type) or one (within-type). `subsetData()` drops
all other cells, so every downstream step — PCA, kernels, skrCCA — works on a
much smaller problem. Do it before `computePCA()`.

```{r subset}
obj <- subsetData(obj, cellTypesOfInterest = c("Tumor", "Fibroblast"))
```

## 3. PCA: use a small `nPCA` for targeted panels

For targeted imaging panels (Xenium is typically 300–500 genes; MERFISH
300–1,000), use **`nPCA = 10–15`**. Too many PCs on a small panel add unstable
noise dimensions that hurt gene-weight reproducibility and score transfer,
without improving cell scores. (For full-transcriptome scRNA-seq, use 20–30.)
The default `nPCA = 40` is intended for large panels; lower it for Xenium.

When `normalizedData` is a BPCells `IterableMatrix`, `computePCA()`
automatically runs `BPCells::svds()`, so the PCA step is itself out-of-core.

```{r pca}
obj <- computePCA(obj, nPCA = 15, center = TRUE, scale. = TRUE)
```

## 4. Build kernels the sparse way — skip `computeDistance()`

The classic pipeline is `computeDistance()` → `computeKernelMatrix()`, which
forms a dense `n x n` distance matrix. For 100,000 cells that is ~80 GB per
block, which is infeasible.

The Gaussian kernel `exp(-0.5 (d / sigma)^2)` is essentially zero once
`d > sigma * sqrt(-2 log(lowerLimit))`. The **sparse path** enumerates only the
cell pairs inside that radius, using a grid-based fixed-radius neighbor search,
and never forms the dense matrix. The result is numerically identical to the
dense path (every entry beyond the radius is zero anyway).

You get the sparse path in one of two equivalent ways:

```{r sparse-kernel}
# Option A (recommended): let computeKernelMatrix choose.
# method = "auto" (the default) picks the sparse path when any cell-type block
# reaches autoThreshold (5000) cells, or the aggregate dense workload reaches
# autoThreshold^2 entries. dropDistances = TRUE (default) frees the @distances
# slot afterward. NOTE: do NOT call computeDistance() first on large data — the
# sparse path builds kernels directly from coordinates and would ignore it.
obj <- computeKernelMatrix(
  obj,
  sigmaValues   = c(0.05, 0.1, 0.2),  # required; no default is derived here
  method        = "auto",             # or force "sparse"
  dropDistances = TRUE
)

# Option B (explicit): call the fused sparse constructor directly.
obj <- computeSparseKernel(obj, sigmaValues = c(0.05, 0.1, 0.2))
```

Notes:

- **Force it** with `method = "sparse"` if you want the sparse path regardless
  of size, or raise/lower `autoThreshold` to move the automatic cutoff. Note
  that `method = "auto"` only skips `computeDistance()` when it actually selects
  the sparse path; if a block falls below the threshold it picks `"dense"` and
  then needs distances. On a machine where the dense matrix would not fit, pass
  `method = "sparse"` explicitly so the choice never depends on the threshold.
- The sparse path supports `distType = "Euclidean2D"` and `"Euclidean3D"`
  (inferred from a `z` coordinate column). `"Morphology-Aware"` distances
  require the dense path.
- `sigmaValues` **must** be supplied — because no distance matrix is built,
  there is nothing to derive a default bandwidth from.
- All other kernel options (`lowerLimit`, `upperQuantile`, `normalizeKernel`,
  `minAveCellNeighor`, distance normalization) behave exactly as in the dense
  path.

## 5. skrCCA is already exact and fast for two cell types

For the ordinary two-cell-type (or single within-type) problem, `runSkrCCA()`
now solves **all** requested canonical axes with one exact SVD of the small
PC-space cross-operator (`nPCA x nPCA`). There is no power iteration to tune, so
`maxIter`, `tol`, and `step_size` do not affect the two-type result — the cost
is dominated by the kernel, not the optimization.

```{r skrcca}
obj <- runSkrCCA(obj, scalePCs = TRUE, nCC = 2)
obj <- computeNormalizedCorrelation(obj)

# Prefer regression gene weights for downstream use / transfer:
obj <- computeRegressionGeneScores(obj)
```

(`maxIter` / `step_size` still matter only when analyzing **three or more**
cell types jointly, or when supplying a transferred first axis, where the
sequential deflation path is used.)

## 6. Multi-slide at scale

`newCoProMulti()` shares one set of weight vectors across slides. Kernels are
built **per slide** and their small PC-space operators are summed — CoPro never
materializes a block-diagonal `N x N` kernel across the whole cohort. Use
`method = "auto"`/`"sparse"` exactly as above, and set `n_cores` to
parallelize the per-slide kernel and skrCCA work.

```{r multi-slide}
multi <- newCoProMulti(
  normalizedData = expr_list,   # named list of per-slide cells x genes matrices
  locationData   = loc_list,
  metaData       = meta_list,
  cellTypes      = ct_list
)
multi <- subsetData(multi, cellTypesOfInterest = c("Tumor", "Fibroblast"))
multi <- computePCA(multi, nPCA = 15, center = TRUE, scale. = TRUE)
multi <- computeKernelMatrix(multi, sigmaValues = c(0.05, 0.1, 0.2),
                             method = "auto", dropDistances = TRUE)
multi <- runSkrCCA(multi, scalePCs = TRUE, nCC = 2, n_cores = 4)
multi <- computeNormalizedCorrelation(multi)
multi <- computeRegressionGeneScores(multi)
```

## 7. A complete single-slide Xenium recipe

```{r recipe}
library(CoPro)

obj <- newCoProSingle(
  normalizedData = xenium_expr,   # dgCMatrix or BPCells IterableMatrix
  locationData   = xenium_loc,    # x, y columns
  metaData       = xenium_meta,
  cellTypes      = xenium_types
)
obj <- subsetData(obj, cellTypesOfInterest = c("Tumor", "Fibroblast"))
obj <- computePCA(obj, nPCA = 15, center = TRUE, scale. = TRUE)

# No computeDistance(): go straight to sparse kernels.
obj <- computeKernelMatrix(obj, sigmaValues = c(0.05, 0.1, 0.2),
                           method = "auto", dropDistances = TRUE)

obj <- runSkrCCA(obj, scalePCs = TRUE, nCC = 2)
obj <- computeNormalizedCorrelation(obj)
obj <- computeRegressionGeneScores(obj)

cell_scores <- getCellScores(obj, sigma = obj@sigmaValueChoice,
                             cellType = "Tumor")
```

## Quick reference: memory-saving knobs

| Setting | Where | Effect |
|---|---|---|
| sparse `dgCMatrix` / BPCells input | `newCoProSingle()` / `newCoProMulti()` | Avoids a dense expression matrix; BPCells keeps it on disk |
| `subsetData()` before PCA | pipeline order | Shrinks every later step to the cell types of interest |
| `nPCA = 10–15` | `computePCA()` | Fewer, stabler dimensions for targeted panels |
| `method = "auto"` / `"sparse"` | `computeKernelMatrix()` | Never forms the dense `n x n` kernel; exact and lossless |
| **skip `computeDistance()`** | pipeline order | The dense distance matrix is the main blow-up; the sparse path does not need it |
| `dropDistances = TRUE` | `computeKernelMatrix()` | Clears `@distances` after kernels are built (default) |
| `autoThreshold` | `computeKernelMatrix()` | Tunes the cell-count cutoff for the sparse path (default 5000) |
| `n_cores` | `runSkrCCA()` (multi-slide) | Parallelizes per-slide kernel and skrCCA work |

## Verifying equivalence (optional)

On a dataset small enough to run both, you can confirm the sparse and dense
paths agree. The stored kernels differ only in representation (`dgCMatrix` vs
dense) and by floating-point rounding:

```{r verify}
# The dense path needs computeDistance() first; the sparse path does not.
obj_dense  <- computeKernelMatrix(computeDistance(obj), sigmaValues = 0.1,
                                  method = "dense", dropDistances = FALSE)
obj_sparse <- computeKernelMatrix(obj, sigmaValues = 0.1, method = "sparse")

k_dense  <- getKernelMatrix(obj_dense,  sigma = 0.1,
                            cellType1 = "Tumor", cellType2 = "Fibroblast")
k_sparse <- as.matrix(getKernelMatrix(obj_sparse, sigma = 0.1,
                            cellType1 = "Tumor", cellType2 = "Fibroblast"))
max(abs(k_dense - k_sparse))   # ~1e-15
```

## See also

- `vignette("brain_merfish_two_type")` — cross-cell-type co-progression on a
  MERFISH panel.
- `vignette("organoid_one_type")` — within-cell-type spatial patterns.
- `?computeKernelMatrix`, `?computeSparseKernel` — full kernel options.
