---
title: "Handling very large datasets (Xenium, large MERFISH)"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Handling very large datasets (Xenium, large MERFISH)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# This is a how-to guide: the code shows the recommended calls for
# datasets that are too large to ship in a vignette, so chunks are not
# executed. Every call below is a valid, current CoPro API call.
#
# Two separate options are needed, and each covers a step the other does not:
#
#   eval = FALSE (here)  -- stops the code running when the vignette is KNITTED.
#   purl = FALSE (on every chunk header below)
#                        -- stops the code being extracted into the tangled
#                           large_datasets.R, which `R CMD check` sources in its
#                           "running R code from vignettes" step.
#
# Without the per-chunk purl = FALSE, `R CMD check` fails with
# "object 'sparse_expr' not found" even though nothing is evaluated at knit
# time. Setting purl through opts_chunk$set() here does NOT work: this chunk is
# not evaluated during tangling, so the option never takes effect. The same
# applies to the YAML `knitr: opts_chunk:` route. It has to be on each chunk
# header.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

Imaging-based platforms such as **10x Xenium** and large **MERFISH** panels
routinely produce hundreds of thousands to millions of cells per slide. Two
things blow up with the number of cells `n`:

1. **The expression matrix** (`n` cells x genes).
2. **The spatial distance / kernel matrix**, which is `n x n` per
   cell-type block if built densely — the dominant cost at scale.

CoPro handles each with a memory-efficient path. This guide collects the
settings that keep a large run within memory while producing results that are
numerically identical to the classic dense pipeline. The scientific
interpretation of the output (cell scores, gene scores, normalized
correlation) is unchanged; only how they are computed differs.

The single most important change for large data is: **do not call
`computeDistance()` — use the sparse kernel path instead.**

## 1. Store expression sparsely or on disk

`newCoProSingle()` and `newCoProMulti()` accept three matrix types for
`normalizedData`:

- a base `matrix` (dense),
- a **sparse** `Matrix::dgCMatrix` (recommended for imaging panels, which are
  mostly zeros), or
- a **BPCells** `IterableMatrix`, which keeps the counts on disk and streams
  them (recommended when the matrix does not fit in RAM).

```{r input-types, purl=FALSE}
library(CoPro)

# Sparse in-memory matrix (genes are columns, cells are rows):
obj <- newCoProSingle(
  normalizedData = sparse_expr,     # a cells x genes dgCMatrix
  locationData   = loc_df,          # data.frame with x, y (and optional z)
  metaData       = meta_df,
  cellTypes      = cell_type_labels
)

# On-disk with BPCells (install with remotes::install_github("bnprks/BPCells/r")):
# `bp_mat` is a BPCells IterableMatrix; computePCA() below will call
# BPCells::svds() so the PCA never materializes a dense matrix.
obj <- newCoProSingle(
  normalizedData = bp_mat,
  locationData   = loc_df,
  metaData       = meta_df,
  cellTypes      = cell_type_labels
)
```

## 2. Subset to the cell types of interest early

A Xenium slide may contain dozens of cell types, but a co-progression analysis
usually targets two (cross-type) or one (within-type). `subsetData()` drops
all other cells, so every downstream step — PCA, kernels, skrCCA — works on a
much smaller problem. Do it before `computePCA()`.

```{r subset, purl=FALSE}
obj <- subsetData(obj, cellTypesOfInterest = c("Tumor", "Fibroblast"))
```

## 3. PCA: use a small `nPCA` for targeted panels

For targeted imaging panels (Xenium is typically 300–500 genes; MERFISH
300–1,000), use **`nPCA = 10–15`**. Too many PCs on a small panel add unstable
noise dimensions that hurt gene-weight reproducibility and score transfer,
without improving cell scores. (For full-transcriptome scRNA-seq, use 20–30.)
The default `nPCA = 40` is intended for large panels; lower it for Xenium.

When `normalizedData` is a BPCells `IterableMatrix`, `computePCA()`
automatically runs `BPCells::svds()`, so the PCA step is itself out-of-core.

```{r pca, purl=FALSE}
obj <- computePCA(obj, nPCA = 15, center = TRUE, scale. = TRUE)
```

## 4. Build kernels the sparse way — skip `computeDistance()`

The classic pipeline is `computeDistance()` → `computeKernelMatrix()`, which
forms a dense `n x n` distance matrix. For 100,000 cells that is ~80 GB per
block, which is infeasible.

The Gaussian kernel `exp(-0.5 (d / sigma)^2)` is essentially zero once
`d > sigma * sqrt(-2 log(lowerLimit))`. The **sparse path** enumerates only the
cell pairs inside that radius, using a grid-based fixed-radius neighbor search,
and never forms the dense matrix. The result is numerically identical to the
dense path (every entry beyond the radius is zero anyway).

First pick sigma from the data with `detectSigmaRange()`, then build:

```{r sparse-kernel, purl=FALSE}
# Step 1: choose sigma in this dataset's own coordinate units. sigma is a
# distance, so no fixed value travels between microns, pixels and Visium units.
# detectSigmaRange() reports the sigmas at which the median cell is coupled to
# 5-20 effective neighbors -- a dimensionless, biologically meaningful target.
rng <- detectSigmaRange(obj, minNeighbors = 5, maxNeighbors = 20)
rng                      # per-block diagnostics and the recommended grid

# Step 2: build. method = "auto" (default) picks the float32 sparse route for
# anything large and reports the density it predicts. dropDistances = TRUE
# (default) frees @distances afterward. No computeDistance() call is needed --
# the sparse routes build kernels directly from coordinates.
obj <- computeKernelMatrix(
  obj,
  sigmaValues   = rng$sigmaValues,
  method        = "auto",
  dropDistances = TRUE
)
```

Notes:

- **Sigma controls memory, steeply.** The search radius grows linearly in sigma,
  so the number of retained pairs grows as `sigma^d` and saturates: past roughly
  two-thirds density a sparse kernel costs *more* than a dense one (12 bytes per
  stored entry against 8). `method = "auto"` predicts the density before
  building and warns, with the sigma that would bring it back under control. If
  you see that warning, the fix is a smaller sigma, not more RAM.
- **Build one sigma at a time** when memory is tight. Kernels for every
  requested sigma are held simultaneously, so a three-sigma grid costs roughly
  three times one sigma. Looping the sigma outside `computeKernelMatrix()` and
  running `runSkrCCA()` inside the loop bounds the peak.
- **Force a representation** with `method = "float32"` (8 bytes per entry,
  streams one block at a time) or `method = "sparse"` (float64, 12 bytes per
  entry, keeps every block's neighbor list; use it for exactness checks). The
  two agree to about 1e-6 in the normalized correlation.
- The sparse path supports `distType = "Euclidean2D"` and `"Euclidean3D"`
  (inferred from a `z` coordinate column). `"Morphology-Aware"` distances
  require the dense path.
- `sigmaValues` **must** be supplied — because no distance matrix is built,
  there is nothing to derive a default bandwidth from.
- All other kernel options (`lowerLimit`, `upperQuantile`, `normalizeKernel`,
  `minAveCellNeighor`, distance normalization) behave exactly as in the dense
  path.

## 5. skrCCA is already exact and fast for two cell types

For the ordinary two-cell-type (or single within-type) problem, `runSkrCCA()`
now solves **all** requested canonical axes with one exact SVD of the small
PC-space cross-operator (`nPCA x nPCA`). There is no power iteration to tune, so
`maxIter`, `tol`, and `step_size` do not affect the two-type result — the cost
is dominated by the kernel, not the optimization.

```{r skrcca, purl=FALSE}
obj <- runSkrCCA(obj, scalePCs = TRUE, nCC = 2)
obj <- computeNormalizedCorrelation(obj)

# Prefer regression gene weights for downstream use / transfer:
obj <- computeRegressionGeneScores(obj)
```

(`maxIter` / `step_size` still matter only when analyzing **three or more**
cell types jointly, or when supplying a transferred first axis, where the
sequential deflation path is used.)

## 6. Multi-slide at scale

`newCoProMulti()` shares one PCA loading matrix and one set of CCA weight
vectors across slides. By default, `computePCA()` centers and scales each
`(slide, cell type)` gene matrix before stacking the blocks for that shared
PCA. `runSkrCCA()` then uses equal-slide SUMCOR, keeping the small per-slide
PC-space operators separate so each cross term is divided by its own score
scales. CoPro never materializes a block-diagonal `N x N` kernel across the
cohort. Use `method = "auto"`/`"sparse"` exactly as above, and set `n_cores`
to parallelize the per-slide kernel work.

```{r multi-slide, purl=FALSE}
multi <- newCoProMulti(
  normalizedData = expr_list,   # named list of per-slide cells x genes matrices
  locationData   = loc_list,
  metaData       = meta_list,
  cellTypes      = ct_list
)
multi <- subsetData(multi, cellTypesOfInterest = c("Tumor", "Fibroblast"))
multi <- computePCA(multi, nPCA = 15, center = TRUE, scale. = TRUE)
multi <- computeKernelMatrix(multi, sigmaValues = c(0.05, 0.1, 0.2),
                             method = "auto", dropDistances = TRUE)
multi <- runSkrCCA(multi, scalePCs = TRUE, nCC = 2, n_cores = 4)
multi <- computeNormalizedCorrelation(multi)
multi <- computeRegressionGeneScores(multi)
```

## 7. A complete single-slide Xenium recipe

```{r recipe, purl=FALSE}
library(CoPro)

obj <- newCoProSingle(
  normalizedData = xenium_expr,   # dgCMatrix or BPCells IterableMatrix
  locationData   = xenium_loc,    # x, y columns
  metaData       = xenium_meta,
  cellTypes      = xenium_types
)
obj <- subsetData(obj, cellTypesOfInterest = c("Tumor", "Fibroblast"))
obj <- computePCA(obj, nPCA = 15, center = TRUE, scale. = TRUE)

# No computeDistance(): go straight to sparse kernels.
obj <- computeKernelMatrix(obj, sigmaValues = c(0.05, 0.1, 0.2),
                           method = "auto", dropDistances = TRUE)

obj <- runSkrCCA(obj, scalePCs = TRUE, nCC = 2)
obj <- computeNormalizedCorrelation(obj)
obj <- computeRegressionGeneScores(obj)

cell_scores <- getCellScores(obj, sigma = obj@sigmaValueChoice,
                             cellType = "Tumor")
```

## Quick reference: memory-saving knobs

| Setting | Where | Effect |
|---|---|---|
| sparse `dgCMatrix` / BPCells input | `newCoProSingle()` / `newCoProMulti()` | Avoids a dense expression matrix; BPCells keeps it on disk |
| `subsetData()` before PCA | pipeline order | Shrinks every later step to the cell types of interest |
| `nPCA = 10–15` | `computePCA()` | Fewer, stabler dimensions for targeted panels |
| `method = "auto"` / `"sparse"` | `computeKernelMatrix()` | Never forms the dense `n x n` kernel; exact and lossless |
| **skip `computeDistance()`** | pipeline order | The dense distance matrix is the main blow-up; the sparse path does not need it |
| `dropDistances = TRUE` | `computeKernelMatrix()` | Clears `@distances` after kernels are built (default) |
| `autoThreshold` | `computeKernelMatrix()` | Tunes the cell-count cutoff for the sparse path (default 5000) |
| `denseThreshold` | `computeKernelMatrix()` | Predicted density above which `"auto"` warns that sparse saves nothing (default 0.3) |
| `detectSigmaRange()` | before kernels | Picks sigma from the data, in its own coordinate units |
| `n_cores` | `runSkrCCA()` (multi-slide) | Parallelizes per-slide kernel and skrCCA work |

## Verifying equivalence (optional)

On a dataset small enough to run both, you can confirm the sparse and dense
paths agree. The stored kernels differ only in representation (`dgCMatrix` vs
dense) and by floating-point rounding:

```{r verify, purl=FALSE}
# The dense path needs computeDistance() first; the sparse path does not.
obj_dense  <- computeKernelMatrix(computeDistance(obj), sigmaValues = 0.1,
                                  method = "dense", dropDistances = FALSE)
obj_sparse <- computeKernelMatrix(obj, sigmaValues = 0.1, method = "sparse")

k_dense  <- getKernelMatrix(obj_dense,  sigma = 0.1,
                            cellType1 = "Tumor", cellType2 = "Fibroblast")
k_sparse <- as.matrix(getKernelMatrix(obj_sparse, sigma = 0.1,
                            cellType1 = "Tumor", cellType2 = "Fibroblast"))
max(abs(k_dense - k_sparse))   # ~1e-15
```

## See also

- `vignette("brain_merfish_two_type")` — cross-cell-type co-progression on a
  MERFISH panel.
- `vignette("organoid_one_type")` — within-cell-type spatial patterns.
- `?computeKernelMatrix`, `?computeSparseKernel` — full kernel options.
