---
title: "Recommended multi-sample workflow: PC-space SUMCOR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recommended multi-sample workflow: PC-space SUMCOR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

CoPro learns shared co-progression directions from several spatial samples.
The recommended fitting route uses **PC-space SUMCOR**: standardize genes
within each slide and cell type, fit one shared PCA per cell type, then
optimize the sum of per-slide standardized spatial associations with the full
ratio gradient. These are already the `CoProMulti` defaults. The code below
spells them out and shows how to inspect numerical diagnostics.

This walkthrough uses the three-slide Colon Day 0 dataset. Code is shown
without execution during package builds; run it locally after downloading
the example data. The existing [gene-space colon example](colon_d0_multi_slide.html)
illustrates the alternative surrogate solver and its spatial plots.

## Prepare the multi-slide object

```{r data}
library(CoPro)
dat <- readRDS(copro_download_data("colon_d0_multi"))
cell_types <- c("Epithelial", "Fibroblast", "Immune")
obj <- newCoProMulti(
  normalizedData = dat$normalizedData,
  locationData = dat$locationData,
  metaData = dat$metaData,
  cellTypes = dat$cellTypes,
  slideID = dat$slideID
)
obj <- subsetData(obj, cellTypesOfInterest = cell_types)
obj <- computePCA(obj, nPCA = 15, center_per_slide = TRUE)
```

Within-slide centering and scaling happen before the shared PCA. The feature
coordinates are therefore shared standardized-gene coordinates. Retaining 15
PCs is an illustrative choice for this targeted panel; the retained space
is part of the model, and changing it can change the fitted directions.

## Construct kernels and fit

```{r fit}
sigma_range <- detectSigmaRange(obj)
sigma_values <- sigma_range$sigmaValues
# One geometry-based bandwidth for this example, in locationData units.
sigma_choice <- sigma_values[ceiling(length(sigma_values) / 2)]
obj <- computeKernelMatrix(
  obj, sigmaValues = sigma_choice, normalizeKernel = TRUE
)
obj <- runSkrCCA(
  obj, space = "pca", objective = "sumcor", slideWeight = "equal",
  nCC = 2, maxIter = 1000, tol = 1e-5
)
getCCAObjective(obj)
```

Kernels connect cells within the same slide. `slideWeight = "equal"` assigns
equal nominal coefficients to slide/pair terms. It does not remove differences
in kernel geometry or guarantee equal total weight per specimen when a specimen
has more sections. Record the slide-to-specimen mapping and the intended
weighting when describing a multi-sample analysis.

The fitting term is `a' K b / (||a|| ||b||)`. It is invariant to score
rescaling, but need not lie between -1 and 1. `normalizeKernel = TRUE`
uses median eligible row-sum scaling when row/column normalization is not
selected; it is neither spectral nor total-mass normalization. Its setting
is part of the fitted objective. The bandwidth above is chosen from geometry,
not claimed to maximize association.

## Inspect convergence and conditioning

```{r diagnostics}
diagnostics <- getCCADiagnostics(obj, sigma = sigma_choice)
diagnostics$components
diagnostics$conditioning
diagnostics$score_norms
diagnostics$objective_traces$CC1
```

`components` reports the solver, stopping status, final objective, projected
gradient norm, tolerance, iterations, and denominator-floor use for each axis.
For a full-gradient fit, `converged = TRUE` means the projected gradient is
within the requested tolerance. `max_iter` or `line_search_stalled` records
why a fit stopped without meeting it; inspect the residual before interpreting
the axis as numerically stationary. A larger iteration budget may help an
iteration-limited fit, but does not cure poor conditioning.

`conditioning` reports each slide/type Gram matrix in whitened PC coordinates,
including its smallest eigenvalue, condition number, and numerical rank.
These coordinates are used for diagnostics even with `scalePCs = FALSE`.
Full numerical rank does not ensure good conditioning. `score_norms` reports
the unfloored norm at the returned direction. A floor encounter can also
occur during a rejected line-search trial, as recorded in `components`.

The smooth stationarity argument assumes positive marginal metrics on the
retained space. A hard denominator floor does not supply that assumption.
If a Gram matrix is singular or poorly conditioned, review the retained PC
space and slide adequacy and document any change. CoPro does not silently add
a ridge through this diagnostic API.

The guarantee is first-order stationarity under the stated conditions. It
does not establish a global maximum or recovery of a population biological
program. Later axes are assessed conditionally on the earlier accepted axes.
An equivalent SUMCOV shortcut is labeled `sumcov_reduction` and checked for
its returned SUMCOR residual. A supplied axis is labeled `supplied`, with
`converged = NA`. Older saved fits have no diagnostic record until refitted.

## Compute scores and regression gene weights

```{r scores}
obj <- computeNormalizedCorrelation(obj, normalizer = "unwhitened")
obj <- computeGeneAndCellScores(obj)
obj <- computeRegressionGeneScores(obj)
epithelial_scores <- getCellScores(
  obj, sigma = sigma_choice, cellType = "Epithelial"
)
```

Use regression gene weights (`geneScoresRegression`) for manuscript gene
rankings and figures. They describe association of genes with the fitted cell
score; changing the fitting space is separate from choosing these reported
gene weights.

The explicit downstream `unwhitened` normalizer divides by the centered
kernel's Frobenius norm as well as the score norms. For centered scores and a
nonzero denominator, this reported statistic is bounded in absolute value by
one. It does not change the directions fitted above. Other downstream
normalizers answer different questions; record which one was used.

## The gene-space alternative

`runGeneSpaceCCA()` works directly in genes and uses its own filtering,
clipping, and per-slide standardization. Under SUMCOR its current iteration
uses frozen denominators. A converged surrogate fixed point need not be a
stationary point of the ratio objective; the Gauss-Seidel sweep does not
remove this limitation. PC-space fitting is the implemented route for the
full-gradient stationarity claim. This distinction does not demonstrate that
PC-space fitting has better biological recovery, since changing spaces also
changes the model.

This walkthrough covers fitting and numerical assessment. Its diagnostics
do not establish statistical significance or independent biological replication.
