---
title: "Within-cell-type spatial patterns (Organoid)"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Within-cell-type spatial patterns (Organoid)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,
                      fig.width = 6, fig.height = 5)
```

## Overview

This vignette demonstrates how CoPro detects **within-cell-type spatial
patterns** using a single cell type. We analyze a 72-hour intestinal
organoid culture imaged by seqFISH, where all cells are epithelial. CoPro
identifies spatially organized gene programs---self-organization patterns
that emerge within a single population.

**What CoPro finds here:** The spatial co-progression of epithelial cells
captures the crypt--villus axis of the organoid---cells along the same
developmental trajectory cluster spatially.

## Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(CoPro)
library(ggplot2)
```

## Download and load data

```{r download-data}
data_path <- copro_download_data("organoid")
dat <- readRDS(data_path)

cat("Cells:", nrow(dat$normalizedData), "\n")
cat("Genes:", ncol(dat$normalizedData), "\n")
```

The dataset contains:

- `normalizedData`: expression matrix (cells x genes), capped at 95th
  percentile, DESeq2 size-factor normalized, log1p-transformed
- `locationData`: spatial coordinates (pixels / 5000)
- `metaData`: cell attributes
- `cellTypes`: all cells labeled as "Epithelial"

## Visualize tissue layout

```{r plot-layout}
plot_df <- data.frame(
  x = dat$locationData$x,
  y = dat$locationData$y
)

ggplot(plot_df, aes(x = x, y = y)) +
  geom_point(color = "steelblue", size = 0.8, alpha = 0.7) +
  coord_fixed() +
  ggtitle("72hr organoid culture, ROI-1") +
  xlab("x") + ylab("y") +
  theme_classic()
```

## Create CoPro object

```{r create-object}
obj <- newCoProSingle(
  normalizedData = dat$normalizedData,
  locationData = dat$locationData,
  metaData = dat$metaData,
  cellTypes = dat$cellTypes
)

obj <- subsetData(obj, cellTypesOfInterest = "Epithelial")
```

## Run the CoPro pipeline

```{r run-pipeline}
# PCA
obj <- computePCA(obj, nPCA = 30, center = TRUE, scale. = TRUE)

# Spatial distance (no normalization for this dataset)
obj <- computeDistance(obj, distType = "Euclidean2D",
                       normalizeDistance = FALSE)

# Test multiple sigma values
sigma_choice <- c(0.01, 0.02, 0.05, 0.1, 0.15, 0.2)
obj <- computeKernelMatrix(obj, sigmaValues = sigma_choice,
                            upperQuantile = 0.85,
                            normalizeKernel = FALSE,
                            lowerLimit = 5e-7)

# Sparse kernel CCA
obj <- runSkrCCA(obj, scalePCs = TRUE, maxIter = 500, nCC = 4)

# Normalized correlation and scores
obj <- computeNormalizedCorrelation(obj, tol = 1e-3)
obj <- computeGeneAndCellScores(obj)
```

## Select optimal sigma

CoPro automatically selects the sigma that maximizes the CC1 normalized
correlation. We visualize the normalized correlation across all sigma
values and canonical components:

```{r plot-ncorr, fig.width=8, fig.height=3.5}
ncorr <- getNormCorr(obj)

ggplot(ncorr, aes(x = sigmaValues, y = normalizedCorrelation)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ CC_index, nrow = 1) +
  xlab("Sigma") +
  ylab("Normalized Correlation") +
  ggtitle("Normalized correlation across sigma values") +
  theme_minimal()
```

```{r select-sigma}
# Use the automatically selected sigma
sigma_opt <- obj@sigmaValueChoice
cat("Selected sigma:", sigma_opt, "\n")
```

## Correlation plot

For a within-type model, the objective is to find cell scores that are
spatially correlated: cells close in space (as measured by the kernel)
should have similar scores. The kernel-smoothed scores (K * scores) vs
raw cell scores should show a positive correlation:

```{r plot-correlation}
df_corr <- getCorrOneType(obj,
  sigmaValueChoice = sigma_opt,
  cellTypeA = "Epithelial",
  ccIndex = 1
)

ggplot(df_corr) +
  geom_point(aes(x = AK, y = B), size = 0.5, alpha = 0.5,
             color = "steelblue") +
  xlab("Epithelial · Kernel (spatially smoothed)") +
  ylab("Epithelial cell scores") +
  ggtitle(paste0("CC1 spatial correlation (sigma = ", sigma_opt, ")")) +
  theme_minimal()
```

## Cell scores in situ

### Continuous scores

```{r plot-insitu, fig.width=6, fig.height=5.5}
cs <- getCellScoresInSitu(obj, sigmaValueChoice = sigma_opt)

# Clamp color scale for contrast
q99 <- quantile(abs(cs$cellScores), 0.99)

ggplot(cs) +
  geom_point(aes(x = x, y = y, color = cellScores), size = 0.8) +
  scale_color_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                        midpoint = 0,
                        limits = c(-q99, q99),
                        oob = scales::squish,
                        name = "Cell score\n(CC1)") +
  coord_fixed() +
  ggtitle(paste0("CC1 cell scores in situ (sigma = ", sigma_opt, ")")) +
  theme_classic() +
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank())
```

### Binarized scores

Binarizing at the median highlights the two spatial groups---high vs
low scoring cells, revealing the organoid's spatial compartments:

```{r plot-binary, fig.width=6, fig.height=5.5}
cs$group <- ifelse(cs$cellScores > median(cs$cellScores), "High", "Low")

ggplot(cs) +
  geom_point(aes(x = x, y = y, color = group), size = 0.8, alpha = 0.8) +
  scale_color_manual(values = c("High" = "#1e17a4", "Low" = "#ffd8d8"),
                     name = "Score group") +
  coord_fixed() +
  ggtitle("Binarized cell scores") +
  theme_classic() +
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank())
```

The spatial pattern reveals the self-organization of the organoid
epithelium---cells are continuously ordered along a spatial gradient
that reflects their position within the organoid structure.

## Top genes associated with spatial progression

Gene scores reflect how strongly each gene contributes to the spatial
co-progression axis:

```{r top-genes, fig.width=5, fig.height=6}
key <- paste0("geneScores|sigma", sigma_opt, "|Epithelial")
gs <- obj@geneScores[[key]][, 1]

top_idx <- head(order(abs(gs), decreasing = TRUE), 20)
top_df <- data.frame(
  gene = factor(names(gs)[top_idx],
                levels = rev(names(gs)[top_idx])),
  score = gs[top_idx]
)
top_df$direction <- ifelse(top_df$score > 0, "positive", "negative")

ggplot(top_df, aes(x = gene, y = score, fill = direction)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("positive" = "tomato",
                                "negative" = "steelblue"),
                    guide = "none") +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  labs(x = NULL, y = "Gene score (CC1)") +
  ggtitle(paste0("Top 20 genes (sigma = ", sigma_opt, ")")) +
  theme_classic()
```

## Data citation

Organoid data from: Heyman Y, Erez M, Burnham P, Nitzan M, Raj A.
*Self-Organization Through Local Cell-Cell Communication Drives
Intestinal Epithelial Zonation.* bioRxiv 2025.11.14.688372;
doi: [10.1101/2025.11.14.688372](https://doi.org/10.1101/2025.11.14.688372)

## Session info

```{r session-info}
sessionInfo()
```
